Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-09-04 letters/neuroimage.tex
Finished letter, minor fixes to the paper

\documentclass{letter}
\usepackage{hyperref}
\signature{Jon Pipitone \\
Kimel Family Translational Imaging-Genetics Laboratory \\
Research Imaging Centre \\
Centre for Addiction and Mental Health \\
Toronto, Ontario}
\address{}
\begin{document}

\begin{letter}{Editorial Board, Neuroimage}
\opening{Dear Sir or Madam:}

@@

In this manuscript we describe a novel automated MRI hippocampal segmentation
algorithm optimised to perform well with only a small number of manually
segmented training images. We believe this is an important contribution since
the expertise and effort needed to perform manual segmentations can be
prohibitive for many clinicians and researchers, and yet when one-of-a-kind or
modifications to existing segmentations need to be made, existing automated
methods require 30-80 training inputs.

Our automated segmentation algorithm, mischievously named MAGeT-Brain, takes
advantage of the neuroanatomical variability that exists in a population that
is being studied in order to automatically identify the hippocampus and its
subfields. Our results demonstrate consistent accuracy in the identification of
the whole hippocampus when compared to "gold-standard" manual segmentations.
Finally, we have made the our algorithm available publically online for use by
other groups, and are pursuing the the contribution of our segmentations to the
Alzheimer's Disease Neuroimaging Initiative image database.

We believe that the techniques we have developed and our findings are a
significant contribution to the neuroimaging community, specifically for those
researchers interested in large scale studies of the hippocampus in the context
of normal brain function and different forms of brain dysfunction.

@@

%\ps{P.S. You can find the full text of GFDL license at}
\encl{Manuscript}

>>>>

\documentclass[12pt]{letter}
\usepackage{hyperref}

% page counting, header/footer
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{graphicx}

\pagestyle{fancy}
\fancyhead[RO,RE]{\includegraphics[width=3cm]{letters/camh_logo.png}}
%\renewcommand{\headheight}{24pt}
%\renewcommand{\footrulewidth}{0.4pt}

\signature{Jon Pipitone \\
Kimel Family Translational Imaging-Genetics Laboratory \\
Research Imaging Centre \\
Centre for Addiction and Mental Health \\
Toronto, Ontario}
\address{}
\begin{document}

\begin{letter}{Editorial Board, Neuroimage}
\opening{Dear Sir or Madam:}
\thispagestyle{fancy}

@@

In this manuscript we describe a novel automated MRI hippocampal segmentation
algorithm optimised to perform well with only a small number of manually
segmented training images. We believe this is an important contribution because
the expertise and effort needed to perform manual segmentation can be
prohibitive for many clinicians and researchers, and yet existing automated
methods generally require between 30 to 80 training segmentations. This
situation makes it infeasible to use these methods for segmentations based on
histological-based digital segmentations (because of their rarity),
high-resolution digital atlases (because of the time needed to segment these
images), or when exploring the effect of variations on a segmentation protocol.

It is for these reasons that we developed the automated segmentation algorithm,
mischievously named MAGeT-Brain, which takes advantage of the neuroanatomical
variability that exists in the target population being studied to bootstrap a
large template library from a small set of manually segmented images. In this
manuscript we rigorously validate this approach on multiple disease
populations, and compare our segmentations with existing popular methods (e.g.
FSL and FreeSurfer).  Our results demonstrate consistent accuracy in the
identification of the whole hippocampus when compared to "gold-standard" manual
segmentations.  Finally, we have made our algorithm available publically online
for use by other groups, and are pursuing the contribution of our segmentations
to the Alzheimer's Disease Neuroimaging Initiative image database.

We believe that the technique we have developed and our findings are a
significant contribution to the neuroimaging community, specifically for those
researchers interested in large scale studies of the hippocampus in the context
of normal brain function and different forms of brain dysfunction.

@@

\encl{Manuscript}

### 2013-09-04 paper.Rnw
Finished letter, minor fixes to the paper

% title and authors
%\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
%Multiple Automatically Generated Templates}
\title{Bootstrapping Multi-atlas Segmentation Using Multiple Automatically
Generated Templates for the Segmentation of the Whole Hippocampus and
Subfields}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,9]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4,5]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,6,9]{Aristotle N. Voineskos}
\author[1,6,8]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health Laboratory, Hospital for Sick
Children, Toronto, ON, Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomaterials and Biomedical Engineering, University of
Toronto, Toronto, ON, Canada}
\affil[8]{Rotman Research Institute, Baycrest, Toronto, ON, Canada}
\affil[9]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
SNT and \freesurfer{} hippocampal volumes for a subset of images were provided by
ADNI, along with quality control data for each \freesurfer{} segmentation
(guidelines described in \citep{Hartig2010}).  Clinical and demographic data for
the entire \adnidataset{} dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb{} was configured with an atlas library composed of the five Winterburn T1
atlases (described in Experiment 2), and size of template library,
registration method, and label fusion method were determined by the optimal
settings found in Experiment 1.  The \freesurfer{} segmentation method was used via
the \verb+run_first_all+ script according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
in the \adnidataset{} dataset were segmented by both methods.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 3, Application of \mb{} to the segmentation    %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 3 Results: Application to the segmentation of first episode
\\ \mb{}ox{schizophrenia patients}}

@@

>>>>

% title and authors
%\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
%Multiple Automatically Generated Templates}
\title{Bootstrapping Multi-atlas Segmentation Using Multiple Automatically
Generated Templates for the Segmentation of the Whole Hippocampus and
Subfields}
\author[1]{Jon Pipitone}
\author[1]{Min Tae Park}
\author[1]{Julie Winterburn}
\author[1,9]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4,5]{Martin Lepage}
\author[1,6,9]{Aristotle N. Voineskos}
\author[1,6,7,8]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health Laboratory, Hospital for Sick
Children, Toronto, ON, Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomaterials and Biomedical Engineering, University of
Toronto, Toronto, ON, Canada}
\affil[8]{Rotman Research Institute, Baycrest, Toronto, ON, Canada}
\affil[9]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
SNT, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).  Clinical and
demographic data for the entire \adnidataset{} dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb{} was configured with an atlas library composed of the five Winterburn T1
atlases (described in Experiment 2), and size of template library,
registration method, and label fusion method were determined by the optimal
settings found in Experiment 1.  The \fsl{} segmentation method was used via
the \verb+run_first_all+ script according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
in the \adnidataset{} dataset were segmented by both methods.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 3, Application of \mb{} to the segmentation  %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 3 Results: Application to the segmentation of first
episode schizophrenia patients}

@@

\section{Acknowledgements}
We wish acknowledge support from the CAMH Foundation, thanks to Michael and
Sonja Koerner, the Kimel Family, and the Paul E. Garfinkel New Investigator
Catalyst Award. MMC is funded by the W. Garfield Weston Foundation and ANV is
funded by the Canadian Institutes of Health Research, Ontario Mental Health
Foundation, NARSAD, and the National Institute of Mental Health (R01MH099167).

Computations were performed on the gpc supercomputer at the SciNet HPC
Consortium. SciNet is funded by: the Canada Foundation for Innovation under the
auspices of Compute Canada; the Government of Ontario; Ontario Research Fund -
Research Excellence; and the University of Toronto

ADNI Acknowledgements: Data collection and sharing for this project was funded
by the Alzheimer's Disease Neuroimaging Initiative (ADNI) (National Institutes
of Health Grant U01 AG024904). ADNI is funded by the National Institute on
Aging, the National Institute of Biomedical Imaging and Bioengineering, and
through generous contributions from the following: Abbott; Alzheimer's
Association; Alzheimer's Drug Discovery Foundation; Amorfix Life Sciences Ltd.;
AstraZeneca; Bayer HealthCare; BioClinica, Inc.; Biogen Idec Inc.;
Bristol-Myers Squibb Company; Eisai Inc.; Elan Pharmaceuticals Inc.; Eli Lilly
and Company; F. Hoffmann-La Roche Ltd and its affiliated company Genentech,
Inc.; GE Healthcare; Innogenetics, N.V.; IXICO Ltd.; Janssen Alzheimer
Immunotherapy Research  Development, LLC.; Johnson \& Johnson Pharmaceutical
Research  Development LLC.; Medpace, Inc.; Merck \& Co., Inc.; Meso Scale
Diagnostics, LLC.; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Servier;
Synarc Inc.; and Takeda Pharmaceutical Company. The Canadian Institutes of
Health Research is providing funds to support ADNI clinical sites in Canada.
Private sector contributions are facilitated by the Foundation for the National
Institutes of Health (www.fnih.org). The grantee organization is the Northern
California Institute for Research and Education, and the study is Rev March 26,
2012 coordinated by the Alzheimer's disease Cooperative Study at the University
of California, San Diego. ADNI data are disseminated by the Laboratory for
NeuroImaging at the University of California, Los Angeles. This research was
also supported by NIH grants P30 AG010129 and K01 AG030514.
